Was Devon Allen Unjustly Disqualified at the 2022 World Track and Field Championships?

### 2023-03-27 Manuscript/manuscript.tex
Added scholarly paper works

After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Allen himself
appeared frustrated and angered with the result. Had he reacted just 0.001 
seconds slower he would have been allowed to compete and been spared waiting
another two years until the next World Track and Field Championship.
%Do I need to cite a video?

Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether or not timing
systems were accurate.  Outside of the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
necessarily suprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as "Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and "The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".
%Should I include how I read LetsRun and talk about any personal experiences
%I have using it?

@@

\section{GLMM Gamma Model}
When determining the most effective model to use, several examples were
explored across the 4 different data sets: a linear model without a year effect,
linear model with year effect, a gamma model without a year effect, a gamma model
with a year effect, a gamma model without a year effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed both
linear models with various distributions to be modeled\citep{Rpkg:lme4}. A GLMM is a type of model
that supports inputs that are from different distributions.  The model that we use is for the
Gamma distribution for the fixed effects part of the model and a random effects part.  In
the model used in the paper, the fixed effect comes in the form of an intercept, and then
there is a random effect which in this case we will call the venue effect.  The
venue effect is responsible for the variation between years and tries to explain
why some years have athletes with faster reaction times than others.

In the code, the 
link parameter under Gamma() was changed to be "log" as logarithmic functions are better 
equipped to deal with high outliers.  The graph below shows the skewed outliers: ~\ref{fig:Skew}.

The effect of adding link = "log" is that

>>>>

1)
After coming in third in the 110 meter hurdles at the United States Track and 
Field Championships in Eugene Oregon at the end of June, Allen was expected to 
be a contender for the World Championships, which were scheduled for two months 
later, also in Eugene. At the World Championships, Allen made it through the 
heats and semifinals to reach the finals while posting reaction times of 0.123
in the heats and 0.101 in the semifinal.  Allen, who attended the University
of Oregon (Location of US World Championship and the World Championship) was
considered to be a favorite for the 110 meter hurdle after running 12.84 seconds
in early June; only 0.04 seconds away from the world record \citep{Preview}.
Thus his disqualification by reacting in 0.099 seconds instead of 0.01 was met
with shock from both Allen and fans alike.  The crowd erupted with jeers when
they found out that Allen was disqualified, angry at the result.  Allen himself
appeared frustrated and angered with the result. Had he reacted just 0.001 
seconds slower he would have been allowed to compete and been spared waiting
another two years until the next World Track and Field Championship.
%Do I need to cite a video?

2) and 3)
Other scholarly papers have discussed track and field reaction times but not in the
context of whether the minimum reaction time barrier is too low or whether timing
systems were accurate.  Outside the blog posts from LetsRun.com, there have not been
any published papers regarding Devon Allen's disqualification in particular.  This is not
necessarily surprising given that his disqualification occurred only last August.
Letsrun is a website that is part message board and part news.  The message board,
which functions similarly to Reddit is very active during important running
events, which include during the finals of major events such as
the Olympics and the World Track and Field Championships.  However, it is the
news section of the website that generated articles such as "Was Devon Allen
Screwed? There's At Least a 99\% That He Was" and "The Data Keeps Pouring In and
It Continues to Look Bad For World Athletics and Great For Devon Allen".  Robert
Johnson, who created and runs the Letsrun website argued that there was an error
with the timing equipment at the 2022 Track and Field World Championship and
cited graphs, statistics such as median reaction time, and a reference to a study
published in cooperation with IAAF in 2009 questioning the validity of the 0.1 
second reaction barrier.  The paper analyzed sprinters reaction times
and the biological processes that went into reacting to the starting gun and
concluded that athletes can react in as low as 0.08 seconds. It even recommended
to IAAF that the disqualification barrier be decreased to allow sprinters to
react faster, and suggested 0.08 or 0.085 seconds as possible new thresholds.
Additionally, it was recommmended to IAAF that high speed cameras be used to
change the reaction time criteria from pushing off on the block to body movement.
This could provide a more accurate representation of reaction time, and thus the
possibility of a false disqualification would be lowered \citep{Komi}.
Despite the study, IAAF has not changed the reaction barrier since it was
implemented at the 1999 World Track and Field Championships.  It is confusing
as to why IAAF would commission a study only to disregard the results and not
make the necessary changes.  It is unknown how many athletes since 2009 could
have potentially benefited from this, but it is probable that there are others
besides Allen who were disqualified.

%Should I include how I read LetsRun and talk about any personal experiences
%I have using it?

4)
This paper will try to model the data using a 
When determining the most effective model to use, several examples were
explored across the 4 different data sets: a linear model without a year effect,
linear model with year effect, a gamma model without a year effect, a gamma model
with a year effect, a gamma model without a year effect and link set to log instead
of the default which is Gaussian.  The R package lme4 was very helpful with this
as it supported the Generalized Linear Mixed Model (GLMM) which allowed both
linear models with various distributions to be modeled\citep{Rpkg:lme4}. A GLMM is a type of model
that supports inputs that are from different distributions.  The model that we use is for the
Gamma distribution for the fixed effects part of the model and a random effects part.  In
the model used in the paper, the fixed effect comes in the form of an intercept, and then
there is a random effect which in this case we will call the venue effect.  The
venue effect is responsible for the variation between years and tries to explain
why some years have athletes with faster reaction times than others.

In the code, the 
link parameter under Gamma() was changed to be "log" as logarithmic functions are better 
equipped to deal with high outliers.

@@

\section{GLMM Gamma Model}
